---
title: "Graph Challenge 5"
author: "Aryah Rao"
date: "`r Sys.Date()`"
format: 
  docx:
    reference-doc: ../quarto_template.docx
    number-sections: true
    number-depth: 3
    fig-dpi: 500
    fig-align: center
    fig-height: 4
    fig-width: 6
execute: 
  echo: true
  warning: false
  message: false
---

The long peace theory generally holds that conflict is less likely now than in the past, but it argues this is true especially for major powers. Create a plot that shows the rate of conflict initiation over time among non-major powers versus major powers to see if this is the case. Use `{peacesciencer}` to create a state-year dataset and then populate it with the variables you need using `add_cow_majors()` and `add_mid_opportunity()` (remember that this is a custom function I made, so check the lecture notes to see how to use it). Use `{geomtextpath}` in some way to help compare trends in a single graph.

First, we create a state-year dataset, add indicators for major powers, and conflict opportunity. This dataset spans from 1816 to 2010.

```{r}
# Load necessary packages
library(tidyverse)
library(peacesciencer)
library(ggplot2)
library(geomtextpath)
source(
  "https://raw.githubusercontent.com/milesdwilliams15/dpr-101-project-files/refs/heads/main/_helper_functions/add_opportunity.R"
)

# Create a state-year dataset and add major power and conflict data
state_year_data <- create_stateyears(subset_years = 1816:2010) %>%
  add_cow_majors() %>%          # Add major power indicator
  add_mid_opportunity()         # Add conflict opportunity

# Summarize data to calculate MID initiation rate for major and non-major powers
data_summary <- state_year_data %>%
  group_by(year, cowmaj) %>%
  summarize(mid_init_rate = sum(mid_inits) / sum(n_pairs), .groups = 'drop') %>%
  mutate(power_status = ifelse(cowmaj == 1, "Major Power", "Non-Major Power"))
```

```{r}
# Find the peaks of each line for labeling
peak_labels <- data_summary %>%
  group_by(power_status) %>%
  filter(mid_init_rate == max(mid_init_rate))

# Plot the conflict initiation rate over time
ggplot(data_summary, aes(x = year, y = mid_init_rate, color = power_status)) +
  geom_line(size = 1.2) +
  geom_text(data = peak_labels, aes(label = power_status, x = year, y = mid_init_rate), hjust = -0.1) +
  labs(
    x = "Year",
    y = "Conflict Initiation Rate",
    title = "Conflict Initiation Rate Over Time:\nMajor vs Non-Major Powers"
  ) +
  theme(legend.position = "none")
```

The plot shows that major powers had significantly higher peaks in conflict initiation rates, especially around World War II, while non-major powers maintained consistently lower rates over time (except the peak around World War II).
